---
title: "Covid91 vaccine study Final2024"
author: "Michael Recinos"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

### Barplot

```{r}
Males <-filter(FinalDat, sex=="M")
```


```{r}
Males <-select(Males,-sex)
```

```{r}
dd2 <- Males %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The males population is represented in this plot along with its relationship to the COVID-19 infection by using either a placebo or the vaccine. 

The data indicates that among those who received the vaccine, only 26% contracted COVID-19, whereas approximately 74% of those given a placebo developed the disease. This suggests that the vaccine significantly reduces the risk of contracting COVID-19. In fact, those who received the placebo were about three times more likely to contract COVID-19 compared to those who were vaccinated. Therefore, the vaccine appears to reduce the risk of contracting COVID-19 by about 1.5 times compared to the placebo. Two plots are described. The first plot has an x-axis labeled "infected," indicating the COVID-19 infection status of participants. The y-axis shows the count of individuals either infected or not infected by COVID-19. The second plot focuses exclusively on individuals with COVID-19, as shown by the left two bars of the graph. The x-axis remains labeled by infection status. Unlike the first plot, the y-axis in the second plot displays the percentage of individuals who contracted COVID-19, distinguishing between those who received a vaccine and those who were given a placebo.

### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=Males)
rowPerc(table1)
colPerc(table1)
```

Numerical data shows that while 74% of men who were given a placebo contracted COVID-19, only 26% of those who received the vaccine did. This indicates that the vaccine significantly reduces the likelihood of contracting COVID-19 compared to a placebo. In summary, the vaccine is more effective than the placebo, with the risk of contracting the virus being about twice as high with the placebo.

### Inferential 

```{r}
chisq.test(table1)
chisqtestGC(table1)
Malesfish <- fisher.test(table1)
Malesfish
```


The results from the Fisher's Exact test are crucial to consider, especially given the extremely small p-value of \(2.2 \times 10^{-16}\). This very low figure suggests that the observed outcomes are highly unlikely to have occurred by chance.

The null hypothesis posits that there is no association between sex and COVID-19 mortality. However, the alternative hypothesis, which suggests a relationship between sex and death rates from COVID-19, appears more plausible based on the strong statistical evidence indicating that sex does indeed influence mortality rates.



```{r}
effect <- Malesfish$p.value/(Malesfish$p.value +1)
effectivenessMales = Malesfish$estimate/(Malesfish$estimate +1)*100
print(effectivenessMales)
```

```{r}
low=(Malesfish$conf.int[1]/(Malesfish$conf.int[1] +1))*100
hi=(Malesfish$conf.int[2]/(Malesfish$conf.int[2] +1))*100
options(digits=2)
```
### Effectiveness Confidence Interval `r low`  to `r hi`

The efficacy of the COVID-19 vaccine for men is depicted above. A vaccine is considered effective if its efficacy rate exceeds 50%. Based on the provided confidence interval, ranging from a low of 68.84 to a high of 78.32, we can determine that the vaccine is indeed effective in preventing COVID-19 among men.

The effectiveness of the vaccine for males is further supported by an odds ratio of 74, which suggests a significant protective effect against the virus.


## Females

### Barplot



```{r}
Females <-filter(FinalDat, sex=="F")
```


```{r}
Females <-select(Females,-sex)
```



```{r}
dd2 <- Females %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

This plot illustrates the relationship between COVID-19 infection rates and the administration of a vaccine or placebo among females. The data suggest that individuals who received the placebo were more likely to contract COVID-19 compared to those who were vaccinated. Specifically, only about 35% of those who contracted COVID-19 had received the vaccine, indicating that approximately 65% of infections occurred in the placebo group. This implies that the risk of contracting COVID-19 after receiving the vaccine is about 1:2.

The first plot displays the number of individuals who contracted COVID-19 on the x-axis, labeled "infected," and the count of individuals on the y-axis. This plot shows the total number of individuals who either contracted COVID-19 or did not.

The second plot focuses on those who contracted COVID-19, represented by the two bars on the left. The x-axis categorizes individuals based on their COVID-19 status, while the y-axis now presents the percentage of individuals who contracted COVID-19 after receiving either the vaccine or a placebo, rather than the total number of cases. This plot emphasizes the proportion of infections among vaccinated and placebo groups.

### Numerical Results

```{r}
table1 <- xtabs(~infected + treatment, data=Females)
rowPerc(table1)
colPerc(table1)
```

The data shows that among girls or women, 35% of those who received the vaccine contracted COVID-19, compared to 65% of those who were given a placebo. This suggests that the vaccine significantly reduces the risk of contracting COVID-19 in this group.

The risk of contracting COVID-19 for vaccinated girls is 1:1, indicating that for every vaccinated individual who contracts the virus, there is one who does not. In contrast, the risk of contracting the virus is reduced by half among those who received the vaccine compared to those in the placebo group. This demonstrates the vaccine's effectiveness in protecting against COVID-19.

### Inferential 

```{r}
chisq.test(table1)
chisqtestGC(table1)
Femalesfish <- fisher.test(table1)
Femalesfish
```

The results of Fisher's Exact test are crucial to consider in this analysis. The extremely low p-value of \(2 \times 10^{-8}\) strongly suggests that the observed findings are not due to random chance. Consequently, we can reject the null hypothesis and accept the alternative hypothesis.

The odds ratio calculated from Fisher's Exact Test is 1.9. This indicates that women who were administered the medication had a 0.32 times higher likelihood of contracting COVID-19 compared to those who were given a placebo.



```{r}
effect <- Femalesfish$p.value/(Femalesfish$p.value +1)
effectivenessFemales = Femalesfish$estimate/(Femalesfish$estimate +1)*100
print(effectivenessFemales)
```

```{r}
low=(Femalesfish$conf.int[1]/(Femalesfish$conf.int[1] +1))*100
hi=(Femalesfish$conf.int[2]/(Femalesfish$conf.int[2] +1))*100
options(digits=2)
```

### Effectiveness Confidence Interval `r low`  to `r hi`


The COVID-19 vaccine is highly effective for females, with an efficacy rate exceeding the 65% threshold. Given that the rate is above 70%, which is the standard for vaccine effectiveness, it's clear that this vaccine is very successful in protecting women from contracting COVID-19.



## LGBTQ

### Barplot

```{r}
LGBTQ <-filter(FinalDat, LGBTQ=="gay")
```


```{r}
dd2 <- LGBTQ %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The bar plot indicates that 19% of LGBTQ individuals who received a COVID-19 vaccine contracted the virus, compared to 80% who received a placebo and also contracted it. This suggests that LGBTQ individuals who were given a placebo had a higher likelihood of contracting COVID-19 than those who were vaccinated.

### Numerical Results

```{r}
table1 <- xtabs(~infected + treatment, data=LGBTQ)
rowPerc(table1)
colPerc(table1)
```

The data indicates that 20% of LGBT individuals who were vaccinated contracted COVID-19, compared to 80% of those who received a placebo. This suggests that the placebo offers no protection against COVID-19 for the LGBT community, and its use is not advisable. 

The likelihood of contracting COVID-19 is 3.2 times higher with the placebo than with the vaccine. This demonstrates a significantly increased risk for those who received the placebo compared to those who were vaccinated.

### Inferential 

```{r}
chisq.test(table1)
chisqtestGC(table1)
LGBTQfish <- fisher.test(table1)
LGBTQfish
```

The p-value presented is exceptionally low, at 2e-06, indicating a very small likelihood that the observed results are due to chance. The null hypothesis posits no connection between gender and COVID-19 mortality rates. However, the substantial evidence suggesting that gender does impact mortality rates supports the alternative hypothesis, which proposes a relationship between gender and COVID-19-related deaths.


```{r}
effect <- LGBTQfish$p.value/(LGBTQfish$p.value +1)
effectivenessLGBTQ = LGBTQfish$estimate/(LGBTQfish$estimate +1)*100
print(effectivenessLGBTQ)
```

```{r}
low=(LGBTQfish$conf.int[1]/(LGBTQfish$conf.int[1] +1))*100
hi=(LGBTQfish$conf.int[2]/(LGBTQfish$conf.int[2] +1))*100
options(digits=2)
```

### Effectiveness Confidence Interval `r low`  to `r hi`

The data provided evaluates the effectiveness of the COVID-19 vaccine among the LGBTQ (gay) community. A vaccine is deemed effective if it has an efficacy rate higher than 50%. According to the time frame indicated, it appears that the vaccine does not sufficiently protect LGBTQ (gay) individuals from contracting COVID-19. 

The odds ratio, which approximates vaccine effectiveness, is 81 for the LGBTQ (gay) community. Additionally, the confidence interval for effectiveness among men ranges from a low of 68.99 to a high of 89.87.


## Druggies

### Barplot

```{r}
Drugs <-filter(FinalDat, DrugUser=="no")
```

```{r}
Drugs <-select(Drugs,-DrugUser)
```


```{r}
dd2 <-  Drugs %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The population that responded "Druguser" to questions about drug usage and its connection to contracting COVID-19 while receiving either the vaccine or a placebo is depicted in this figure.

According to this barplot, around 51% of respondents who admitted to drug use and received the vaccination also had COVID-19. According to this, almost 49% of drug users who merely took a placebo ended up with COVID-19.There is a 1:1 chance of contracting COVID-19 while receiving the vaccination. The risk of contracting COVID-19 was halved among drug users who received the vaccine compared to those who received a placebo.

Two plots are present. The x-axis in the first plot indicates infected, denoting whether or not the participants have COVID-19 infection. The count, or total number of individuals who received COVID-19 or not, is shown on the y-axis.

In the above-mentioned second plot, The All of the individuals with COVID-19 are represented by the graph's left two bars. Whether or not a person has COVID-19 is shown on the x-axis. However, the y-axis now shows the percentage of patients who contracted COVID-19 after receiving the vaccine or a placebo, rather than the total number of cases of this disease. 

### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data= Drugs)
rowPerc(table1)
colPerc(table1)
```

Numerical data shows that 33% of drug users who were vaccinated against COVID-19 contracted the virus, compared to 67% of those who received a placebo. These results clearly demonstrate the efficacy of the COVID-19 vaccine in reducing the risk of infection among drug users. It is advisable for drug users to get vaccinated, as the vaccine halves the likelihood of contracting COVID-19 compared to the placebo group. In summary, the vaccine significantly decreases the risk of COVID-19 among drug users, underscoring its protective benefits.

### Inferential 


```{r}
chisq.test(table1)
chisqtestGC(table1)
Drugsfish <- fisher.test(table1)
Drugsfish
```

The p-value from the analysis is 0, suggesting a statistically significant result unlikely due to random chance. According to the Fisher's Exact Test, the odds ratio is 3. This indicates that vaccine recipients had a 40-50% reduced risk of contracting COVID-19 compared to those who received a placebo. Consequently, the null hypothesis is rejected, supporting the effectiveness of the medication in decreasing the risk of COVID-19 among drug users.

```{r}
effect <- Drugsfish$p.value/(Drugsfish$p.value +1)
effectivenessDrugs = Drugsfish$estimate/(Drugsfish$estimate +1)*100
print(effectivenessDrugs)
```

```{r}
low=(Drugsfish$conf.int[1]/(Drugsfish$conf.int[1] +1))*100
hi=(Drugsfish$conf.int[2]/(Drugsfish$conf.int[2] +1))*100
options(digits=2)
```

### Effectiveness Confidence Interval `r low`  to `r hi`

The data presented indicates that the COVID-19 vaccine is effective for men, with an efficacy threshold of over 40% considered sufficient for effectiveness. For drug users, this vaccine demonstrates a high level of effectiveness, exceeding the 60% standard typically required for vaccine efficacy. The effectiveness of the vaccine for drug users is quantified by an odds ratio of 75, with a confidence interval ranging from 71.42 to 78.22. This confirms that the vaccine is highly effective in preventing COVID-19 among drug users during the observed period.

# Overall Results and Conclusions

The COVID-19 Placebo did not provide effective protection against contracting the disease for men (74%), women (65%), and LGBTQ individuals (80%). In contrast, vaccination rates showed varying degrees of effectiveness: 20% for men, 35% for women, and 20% for LGBTQ individuals, indicating a generally lower likelihood of protection. For drug users, both the vaccine (79.7% effectiveness) and the placebo (78.5% effectiveness) were reported to significantly reduce the chances of contracting COVID-19, suggesting that there is not much difference in effectiveness between the two. It is recommended that LGBTQ individuals consider other protective measures as the vaccine does not significantly decrease their risk. Conversely, drug users might benefit from vaccination as it still provides substantial protection.
