---
title: "Covid91 vaccine study Final2024"
author: "Michael Recinos"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

### Barplot

```{r}
Males <-filter(FinalDat, sex=="M")
```


```{r}
Males <-select(Males,-sex)
```

```{r}
dd2 <- Males %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The males population is represented in this plot along with its relationship to the COVID-19 infection by using either a placebo or the vaccine. 

The data indicates that among those who received the vaccine, only 26% contracted COVID-19, whereas approximately 74% of those given a placebo developed the disease. This suggests that the vaccine significantly reduces the risk of contracting COVID-19. In fact, those who received the placebo were about three times more likely to contract COVID-19 compared to those who were vaccinated. Therefore, the vaccine appears to reduce the risk of contracting COVID-19 by about 1.5 times compared to the placebo. Two plots are described. The first plot has an x-axis labeled "infected," indicating the COVID-19 infection status of participants. The y-axis shows the count of individuals either infected or not infected by COVID-19. The second plot focuses exclusively on individuals with COVID-19, as shown by the left two bars of the graph. The x-axis remains labeled by infection status. Unlike the first plot, the y-axis in the second plot displays the percentage of individuals who contracted COVID-19, distinguishing between those who received a vaccine and those who were given a placebo.

### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=Males)
rowPerc(table1)
colPerc(table1)
```

Numerical data shows that while 74% of men who were given a placebo contracted COVID-19, only 26% of those who received the vaccine did. This indicates that the vaccine significantly reduces the likelihood of contracting COVID-19 compared to a placebo. In summary, the vaccine is more effective than the placebo, with the risk of contracting the virus being about twice as high with the placebo.



## Females


## LGBTQ



## Druggies


# Overall Results and Conclusions